---
title: "Whale sightings and El Niño"
format: html
editor: visual
---

During El Niño, California Current upwelling decreases and krill biomass drops. Blue whales feed exclusively on krill, so during these times blue whales shift their distributions farther north.

Here you have two CSV files. One contains a time series of the Oceanic Niño Index (ONI) since 2020. The other has blue whale observations in Southern California during the same time period. You'll use a combination of **pivoting**, **date components**, and **joining** to see if there's a relationship between El Niño and whale sightings.

## Setup

Load the tidyverse and import the two datasets into variables named `oni` and `bluewhales`.

```{r}
# Load the tidyverse

# Replace NULL in the following lines to import the data
oni <- NULL
bluewhales <- NULL

```

```{r}
#| label: test-setup
. = ottr::check("tests/setup.R")
```

## Pivot longer

Inspect `oni`. Notice the months are in columns. Pivot it longer and call the new variable `oni_long`. The result should have columns named `month` and `oni`.

After pivoting, the R will think the month column is text (because it used to be column names). The call to `mutate()` I've given you converts months from text to numbers.

::: callout-tip
You can pivot all columns *except* one using minus (`-`) before the column you want to exclude.
:::

```{r}
# Replace NULL in the following lines to pivot oni longer
oni_long <- NULL |>
  mutate(month = as.numeric(month))

```

```{r}
#| label: test-ex1
. = ottr::check("tests/ex1.R")
```

## Extract date components

We want to be able to match ONI with monthly whale sightings. To do that, we'll have to:

1.  Extract year and month from the datetimes in `bluewhales`.
2.  Group by year and month, then summarize by the count using `n()`.

Create a new variable called `monthly_whales`. Use `year()` and `month()` to extract those components from the `eventDate` column in `bluewhales`. Call these columns `year` and `month`. Then group by both those columns, and summarize using `n()` to create a summary column called `n_whales`.

::: callout-note
You're using multiple grouping columns here, so make sure to use `.groups = "drop"` in `summarize()`.
:::

```{r}
# Replace the NULLs in the lines below to create monthly_whales according to the
# directions above.
monthly_whales <- bluewhales |>
  mutate(NULL) |>
  group_by(NULL) |>
  summarize(NULL)

```

```{r}
#| label: test-ex2
. = ottr::check("tests/ex2.R")
```

## Join

Now you have two data frames, `oni_long` and `monthly_whales`, with both year and month columns. These shared columns are the *keys* for a join. Join the two data frames together on both keys. Your resulting data frame should have columns `year`, `month`, `oni`, and `n_whales`. Call it `oni_whales`.

```{r}
# Replace NULLs in the following lines to create oni_whales
oni_whales <- left_join(NULL,
                        NULL,
                        by = NULL)
```

```{r}
#| label: test-ex3
. = ottr::check("tests/ex3.R")
```

## Visualize

Create a scatter plot of ONI vs the whale count.

```{r}
# Use ggplot to create a scatter plot of the two species.

```

```{r}
#| label: test-ex4
. = ottr::check("tests/ex4.R")
```

**Q1: In one sentence, describe the relationship between ONI and whale counts depicted in the figure.**

*Answer Q1 here.*

## Submission

Download whales-oni.qmd and submit it on Gradescope.
